Writing Trash Bin

### 2024-03-21 paper.qmd
finished decision making section

#### Decision making and risk aversion
Decision making tasks are often described by authors as a more realistic version how plots are used in practice. While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained. While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environement for evaluating plots. Let us consider this in more detail. A normal value extraction task involves:
1) correctly interpret the question
2) extract the specified value
3) report the value
On the other hand, a decision making task involves:
1) Correctly interpret the question
  2) Make a choice that involves
  - If no ground truth is in the experiment
      a) using *personal* risk utility function to identify a threshold
      b) extracting that threshold value from the plot
      c) making a judgement based on that value
  - If a group truth is specified (through right/wrong answers or a payment sceme)
      a) Use the payment sceme to work out the choice threshold that will maximise payout
      b) Over-ride the personal risk utility function adopt the experimenters risk utility function to identify a threshold
      c) extracting that threshold value from the plot
      d) making a judgement based on that value
The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the

Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

Risk and uncertainty are not only different in how they are communicated, but also why they are communicated. Risk communication focuses on communicating probability around events. These events are typically a binary event that is known to be random, so a version of communication that does not mention risks (e.g. telling someone they will or will not get cancer instead of communicating their risks about it) inherently carries with it a lack of transparency. Additionally risk communication is often communicated to avoid or information about an unfavorable outcome, so framing can be important if we want to influence decisions in a certainty way. These motivations do not extend to the communication of uncertainty information more broadly.

There are some elements of risk communication that seem to translate to uncertaitny communication more broadly. Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated, however this can make information harder to understand for those with poor numeracy skills [@Spiegelhalter2017]. People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly *inc (citations for this)*. Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017]. These general concepts of communication also extend to communication of uncertainty. On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication. In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.

In the same sense that risk and uncertainty are different, risk-aversion and amiguity-aversion (which can be considered the uncertainty version of risk aversion) are also not the same. Risk-aversion is the tendency of people to prefer outcomes of low uncertainty to those outcomes with high uncertainty despite the outcomes with high uncertainty having a higher or equal expected outcome. In the case of communicating risk, the only reason this heuristic would be relevant was if risk-aversion was a mistake that needed to somehow be corrected for, rather than a heuristic that *reveals the utility of decreased uncertainty*. Risk-aversion is often treated as a mistake or something that should be avoided by the uncertainty visualisation literature even though that is not necessarily true, depending on what is causing the risk-aversion (e.g. is it due to poorer estimates or increased awareness of negative outcomes). If a particular communication method leads people to place a greater value on certainty or make worse estimates, that is the aspect of importance, not risk-aversion which conflates these factors. The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions. Risk-aversion has a large number of confounding factors that make it difficult to understand if it is something to be considered in risk communication, ambiguity-aversion has a similar problem. Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks. It is essentially risk-aversion applied to risk. For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased uncertainty, it contains a subtext that argues against transparency, and it is considered a valuable insight in of itself despite conflating many factors.

The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue. However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time. If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out. Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*. There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all. The concept of risk aversion treats risk as "noise", as though it is something that only exists to be ignored, but if the common place use of biased MLE's is an indication of anything, it is that *ignoring uncertainty* is the irrational choice.

While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature. Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large. While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept. In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying. Visualisations are used in both data exploration and communication. This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is. Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation. Two people looking at the same visualisation might, just by chance, develop two entirely different insights. These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.

What is meant by "uncertainty" may seem obvious to some, but when you attempt to quantify or visualise it you will quickly find yourself asking, "uncertainty about... what?". Do you mean uncertainty on an estimate? On a forecast? How many steps ahead is this forecast? Are we only considering the uncertainty in the estimate or in the parameters or are we considering the possibility of measurement error or biased inputs? Signal and noise can only be untangled in the presence of a motivating question.

The previous sections of this paper established several issues in uncertainty communication at large. In this section, we will discuss the current literature on uncertainty visualisation, how the uncertainty communication issues more broadly manifest in the specific case of visualisation, and what general lessons we can take from the existing work.

@@

## Suggestions to measure uncertainty
### Heuristic checks
- Have a better hypothesis? Like its more specific. They are looking for a very specific thing *wrong* with a visualisation.
- We wary of trying to control something and observe it at the same time

>>>>

#### Decision making and risk aversion
Decision making tasks are often described by authors as a more realistic version how plots are used in practice. While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained. While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environement for evaluating plots. Let us consider this in more detail. A normal value extraction task involves:

1) correctly interpret the question
2) extract the specified value
3) report the value

On the other hand, a decision making task using an uncertainty visualisation involves:

1) Correctly interpret the question
2) Use some risk utility function to set a threshold for an "acceptable level of risk"
3) Extract the value for from the plot that will be compared to the threshold
4) Make a judgement based off that value

The key aspect that separates a decision making task from a value extraction task, for uncertainty visualisations, is the inclusion of a utility function. While this might seem like a small change, it actually completely warps the experiment in unexpected.

The first issue is that individuals will all have their own individual risk utility function, that is, how much they want to avoid or engage with risks or uncertainty. Therefore decision making experiments have an additional layer of noise that value extraction experiments do not. We cannot be sure if particpants answered differently from the ground truth because a visualisation was difficult to read *or* because the participants risk utility function did not align with the one set by the authors. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however this method seems to just introduce more noise. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*. The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another. Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in sucessful participant responses because it is not in the "spirit" of visualisation.

It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking]. This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017]. In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation. The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this. Asking participants to answer questions that require complicated calculations does not reflect how visualisations are used, and will liekly be ignored by the participants (as seen in many of these decision making tasks). While we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise, we differ in the sense that we do not think it should be done, lest you create a needlessly complicated value extraction task.

Additionally, several papers mention "risk-aversion" in participant decisions as a negative by-product of a particular visualisation design, however that displays a failure to understand that risk aversion is *not* a mistake, but rather it *reveals the utility of decreased uncertainty*. It may be interesting to investigate why a particular visualisation elicits increased risk-aversion, for example is it due to poorer estimates or increased awareness of negative outcomes, however simply reporting this as "risk aversion" conflates these factors. The work that advocates for transparency in risk-communication gives you whiplash when it's subtext argues that transparency leads to "incorrect" conclusions. There have been other discussions on the appropriateness of discussing risk-aversion as a bias [@Vranas2000], but few have made the point that if uncertainty holds some *value* there is not technically a "right" decision at all.

The final issue with decision making tasks is that calculating "optimum" choices is not a task we use uncertainty for, it is a task that utilises *risk*. Risk and uncertainty are slightly different, as risk is known probabilities and uncertainty is unknown probabilities, so uncertainty is what you get when you cannot accurately define risk [@Spiegelhalter2017]. Communicating risk allows people to weigh up options and make an optimum decision, uncertainty hedges information to let people know there may not even be an optimum strategy. In this sense, many "uncertainty" decision making experiments are actually just risk decision making experiments.

This leaves us with several issues in decision making experiments. They are noisy because of participants individual utility functions, and when that is not the case they are thinly veiled and over complicated value extraction experiments, and on top of that, they dont even evaluate uncertainty as uncertainty, they evaluate it as risk. For these reasons, uncertainty visualisation experiments should either be designed with these caveates in mind and be designed to untangle why a particular visualisation will elicit a specific decision, or the field should steer away using decision quality and related metrics as a basis for evaluating visualisation performance.

@@

## Suggestions to measure uncertainty
While uncertainty visualisation can make it easier to communicate the complicated details of uncertainty information, it carries some challenges that are unique to the visualisaition, specifically with respect to the "signal supression" concept. In simple estimates or verbal communication, the signal is often easy to identify because it is what we are explicitly saying. Visualisations are used in both data exploration and communication. This means what exactly is a *signal* in any particular visualisation is hard to identify, since we often let the visualisation *tell us* what the signal is. Additionally, you cannot add noise to *every single possible* signal one might take from a visualisation. Two people looking at the same visualisation might, just by chance, develop two entirely different insights. These unique and facinating challenges that are faced by viewing uncertainty visualisaiton through the lense of "signal supression" have been almost completely untouched by the literature.

### Heuristic checks
- Have a better hypothesis? Like its more specific. They are looking for a very specific thing *wrong* with a visualisation.
- We wary of trying to control something and observe it at the same time

### 2024-03-21 writingdumpster.qmd
finished decision making section

# Main section outtakes
These results provide some insights that applies to visualisation more broadly but nothing specific to uncertainty informaiton. It is clear that audiences primarily interact with visualisations through their "System 1" brain [@daniel2017thinking]. This aligns with other authors comments that visualisation is primarily about "gists" [Spiegelhalter2017]. In this sense, asking participants a question that requires them to shift to their "System 2" is entirely contrary to *why* we use visualisation. The exact reasons we describe visualisation to be so powerful, that they are able to communicate complicated information fast and efficiently *is a product of the system 1 brain* [@daniel2017thinking], however this means displaying information using visualisations have the same weakness' as system 1 thinking, and visualisation authors need to be aware of this. Asking participants to answer questions that require shifting to their system 2 brain is not in the spirit of visualisation, and does not reflect how visualisations will be used in practice, and will very likely result in incorrectly answered questions. 
There is a large number of examples where participants got "confused" but the problem is more likely that the question required participants to shift to their system 2 brain.

>>>>

# Main section outtakes
While this is a focus on uncertainty visualisation, these issues in uncertainty communication need to be acknowledged because they bleed into the uncertainty visualisation literature. Uncertainty visualisation are compared on their ability to communicate explicit risk, elicit trust, and prevent risk or ambiguity avoidance, continuing these issues in uncertainty communication at large.

Ambiguity-aversion, uncertainties risk-avoidance counterpart, has a similar problem. Ambiguity-aversion is the tendency of people to prefer to take on certain risks rather than unknown risks. It is essentially risk-aversion applied to risk. For this reason discussions of ambiguity-aversion have the same issues as the conversations about risk-aversion, where they do not consider ambiguity-aversion to be an estimate of the utility of decreased certainty. Assuming that participants should act as rational agents implies they should do as rational agents do and *ignore* the uncertainty information.

The idea that risk and ambiguity aversion reveal the utility of decreased variance seems to seldom be considered in the literature, therefore it should be mentioned here, otherwise this issue might continue. However, considering risk aversion is at its core, a trade off between bias and variance, it is bizzare to assume it to be a mistake when statisticians perform these trade offs all the time. If risk-aversion is truly "irrational" behaviour, every statistical textbook that discusses the use of a biased but consistent estimator in a finite sample case (a common example would be specific cases of the maximum likelihood estimator) should be tossed out. Discussing risk and ambiguity aversion as though it is some diversion from a perfectly rational, mathematical choice, implies there is *no value to a decrease in variance which is not even true within mathematics*. 
There are some elements of risk communication that seem to translate to uncertaitny communication more broadly. Using words to communicate risk is discouraged because people can misinterpret the actual risk involved, so when possible numerical estimates of risk should be communicated [@Spiegelhalter2017] the same rules apply to uncertainty (*Cite gap __*). 
People performing better and preferring with numerical estimates of risk translates to uncertainty communication more broadly . Risk communication also needs to have clear objectives, use plain language, limit information to only what is necessary, and segment the audience to allow for differences in interest and knowledge [@Spiegelhalter2017]. These general concepts of communication also extend to communication of uncertainty. On the other hand, it is not sensible to assume that probability specific concepts such as framing, intuitiveness of probabilities, or reference classes are still relevant when we consider uncertainty communication as signal supression, not as a synonym of risk or probability communication. In this same vein, most uncertainty communication papers very often discuss risk-aversion as something that needs to be considered despite it being unclear exactly how risk-aversion is related to uncertainty communication *(citations for the papers that do this)*.
Very often the decisions made in decision making tasks are compared to some "correct" choice, typically the choice of a rational agent. That is, visualisations that elicit choices that are most closely aligned with those of rational agents are seen as the best. There is a critical error with this way of thinking. If a group truth is specified (through right/wrong answers or a payment scheme) to correctly answer the question, the user needs to work out the threshold that will maximise payout, over-ride their personal risk utility function to adopt this threshold, and report values that align with this threshold. This is just a value extraction task with additional unnecessary extra steps.
Additionally, some authors speak about plots exhibiting "risk-avoidance" which is comical, considering a rational agent who displays no risk-avoiding will simply make the choice that maximise expected value, or in other words **completely ignores the uncertainty information**. The ground truths in these experiments *encourage* participants to think of uncertainty information as useless noise that should be ignored, the way a rational agent will.
